---
title: "Ordinary least squares"
author: "Gota Morota<br>[<font color=‘white’>http://morotalab.org/</font>](http://morotalab.org/)"
date: "1/27/2020"
output: 
    prettydoc::html_pretty:
      theme: architect
      highlight: vignette
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Background
This exercise illustrates how to perform ordinary least squares (OLS) to identify which SNPs are significantly influencing phenotypes. We will ignore population structure in this example. 


## Install and load R packages
First, we are going to use the `maize` data included in the [synbreedData](https://cran.r-project.org/web/packages/synbreedData/index.html) package. Learn more about the [Synbreed project](http://www.synbreed.tum.de/) and [synbreed R packages](http://synbreed.r-forge.r-project.org/). 
```{r, eval=FALSE, echo=TRUE}
# install.packages("synbreed")
# install.packages("synbreedData")
library(synbreed)
library(synbreedData)
data(maize)
?maize
summary(maize)
```

## Read data
```{r, eval=FALSE, echo=TRUE}
rm(list=ls()) # delete the objects in the current environment 

names(maize)
y1 <- as.vector(maize$pheno) # phenotype vector 
intercept <- rep(1, length(y1)) # intercept vector 
W1a <- cbind(intercept, maize$geno[,1:100]) # the intercept and the SNP matrix including the first 100 SNPs

length(y1) # 12150
dim(W1a) # 1250 101
```

## Fitting OLS using the matrix multiplication operation
```{r, eval=FALSE, echo=TRUE}
?solve # help page for the solve() function
?t # help page for the t() function

fit1a <- solve(t(W1a) %*% W1a) %*% t(W1a) %*% y1
head(fit1a) # estimates of the intercept and the first five SNP effects
```

## lm() function
We will use the `lm()` function to fit OLS and estimate fixed SNP effect coefficients. The summary() function summarizes a model fit.
```{r, eval=FALSE, echo=TRUE}
?lm # help page for the lm() function?

summary(lm(y1 ~ -1 + W1a))
names(summary(lm(y1 ~ -1 + W1a))) # names of accessible lists
summary(lm(y1 ~ -1 + W1a))$coef # return only coefficients
head(summary(lm(y1 ~ -1 + W1a))$coef) # return the first siz coefficients

head(lm.fit(W1a, y1)$coef) # fit the model using the lm.fit() function
```


## t-values 
How to obtain t-values in the third column?  
```{r, eval=FALSE, echo=TRUE}
t0 <- 169.655798 / 16.796230 
t1 <- 2.054498 /  1.362064
t2 <- 8.685180 / 8.942172
t3 <- -7.896607 /  9.480402
t4 <- 4.005036  / 3.790624
t5 <- -3.425773 /  1.569304
```


## p-values 
How to obtain p-values in the fourth column?  
```{r, eval=FALSE, echo=TRUE}
my.df <- 1250 - 1 - 100

2*pt(q=t0, df=my.df, lower=FALSE) # intercept
2*pt(q=t1, df=my.df, lower=FALSE) # M1
2*pt(q=-t2, df=my.df, lower=TRUE) # M2
2*pt(q=t3, df=my.df, lower=FALSE) # M3
2*pt(q=t4, df=my.df, lower=TRUE) # M4
2*pt(q=t5, df=my.df, lower=TRUE) # M5
```



## Fitting the whole markers as fixed
```{r, eval=FALSE, echo=TRUE}
W1b <- cbind(intercept, maize$geno) # the intercept and the whole SNP matrix 
dim(W1b) # 1250 1117

# use the solve() function
fit1b <- solve(t(W1b) %*% W1b) %*% t(W1b) %*% y1

# use the lm() function
summary(lm(y1 ~ -1 + W1b)) # check the warning 
```


## Load the cattle data
```{r, eval=FALSE, echo=TRUE}
data(cattle)
?cattle
dim(cattle$geno)

y2 <- cattle$pheno[,1,1] 
set.seed(100)
cattle2 <- codeGeno(cattle,impute=TRUE,impute.type="random", reference.allele = "minor")
W2 <- cattle2$geno

intercept2 <- rep(1, length(y2)) # intercept vector 
W2 <- cbind(intercept2, cattle2$geno) # the intercept and the SNP matrix 

length(y2) # 500
dim(W2) # 500 7250 + 1
```


## Fit OLS to the cattle data
What happens when we fit OLS to the data with $n < m$?
```{r, eval=FALSE, echo=TRUE}
fit2 <- solve(t(W2) %*% W2) %*% t(W2) %*% y2  

head(summary(lm(y2 ~ -1 + W2))$coef) 
```



