---
title: "Genomic BLUP (GBLUP)"
author: "Gota Morota"
date: "March 27, 2020"
output: 
    prettydoc::html_pretty:
      theme: architect
      highlight: vignette
      toc: true
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


This example illustrates how to fit genomic BLUP (GBLUP) using the [rrBLUP](https://cran.r-project.org/web/packages/rrBLUP/index.html) R package, two-step apporach, and mixed model equations. 

# rrBLUP

## Data 
We will use the wheat data available in the [BGLR](https://cran.r-project.org/web/packages/BGLR/index.html) R package. 
```{r, eval=FALSE}
rm(list=ls())
library(BGLR)
data(wheat)
?wheat
y <- 10 + wheat.Y[,2] # use the second phenotype
W <- wheat.X
```


## Genomic relationship matrix
```{r, eval=FALSE}
p <- colMeans(W)
maf <- ifelse(p > 0.5, 1 - p, p)  # or pmin(p, 1-p)
maf.index <- which(maf < 0.1)
p2 <- p[-maf.index]
W2 <- W[, -maf.index]
dim(W2)

Wcs <- scale(W2, center = TRUE, scale = TRUE)
G <- tcrossprod(Wcs)/ncol(Wcs)
```


## Genomic BLUP
The `mixed.solve()` function implements GBLUP. 
```{r, eval=FALSE}
# install.packages("rrBLUP")
library(rrBLUP)
?mixed.solve
fit <-  mixed.solve(y = y, K = G)

fit$beta # BLUE (intercept)

head(fit$u) # BLUP (genetic values, breeding values)

fit$Vu # additive genetic variance

fit$Ve # residual variance

lambda <- fit$Ve / fit$Vu # lambda
lambda
```




# Two-step approach
The fixed effect incidence matrix $\mathbf{X}$ only includes the intercept. 
```{r eval=FALSE, echo=TRUE}
X <- matrix(1, nrow = nrow(G), ncol=1)
Z <- diag(nrow(G))
I <- diag(nrow(G))
Ginv <- solve(G)
V <- G*fit$Vu + I*fit$Ve
Vinv <- solve(V)

b <- solve(t(X) %*% Vinv %*% X) %*% t(X) %*% Vinv %*% y # BLUE (intercept)
g <- solve(I + Ginv*lambda) %*% (y - X %*% b) # BLUP (genetic values)

b
head(g)
tail(g)
```


# Mixed model equations (MME)
```{r eval=FALSE, echo=TRUE}
XtX <- t(X) %*% X # X'X
XtZ <- t(X) %*% Z # X'Z
ZtX <- t(Z) %*% X # Z'X
ZtZ <- t(Z) %*% Z # Z'Z
Xty <- t(X) %*% y # X'y
Zty <- t(Z) %*% y # Z'y

LHS1 <- cbind(XtX, XtZ)
LHS2 <- cbind(ZtX, ZtZ + Ginv * lambda)
LHS <- rbind(LHS1, LHS2) # left hand side of MME
RHS <- rbind(Xty, Zty) # right hand side of MME

sol <- solve(LHS, RHS) # solve for BLUE and BLUP

# BLUE (intercept)
sol[1, ]

# BLUP (genetic values)
sol[-1, ]
head(sol[-1,])
tail(sol)
```


# Comparison
We will compare the results obtained from the rrBLUP, two-step approach, and MME. Verify that they are all equal. 

Intercept.
```{r eval=FALSE, echo=TRUE}
fit$beta # rrBLUP
b # Two-step approach
sol[1,] # MME
```

The first six estimates of the genetic values. 
```{r eval=FALSE, echo=TRUE}
head(fit$u) # rrBLUP
head(c(g)) # Two-step approach
head(sol[-1,]) # MME
```

The last six estimates of the genetic values. 
```{r eval=FALSE, echo=TRUE}
tail(fit$u) # rrBLUP
tail(c(g)) # Two-step approach
tail(sol[-1,]) # MME
```
