---
title: "Weighted least squares single-marker GWAS"
author: "Gota Morota"
date: "March 27, 2020"
output: 
    prettydoc::html_pretty:
      theme: architect
      highlight: vignette
      toc: true
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


This example illustrates how to fit weighted least squares for single-marker GWAS. 

# Weighted least squares for single-marker GWAS

## Data 
We will use the wheat data available in the [BGLR](https://cran.r-project.org/web/packages/BGLR/index.html) R package. 
```{r, eval=FALSE}
rm(list=ls())
library(BGLR)
data(wheat)
?wheat
y <- 10 + wheat.Y[,2] # use the second phenotype
W <- wheat.X
```


## Genomic relationship matrix
```{r, eval=FALSE}
p <- colMeans(W)
maf <- ifelse(p > 0.5, 1 - p, p)  # or pmin(p, 1-p)
maf.index <- which(maf < 0.1)
p2 <- p[-maf.index]
W2 <- W[, -maf.index]
dim(W2)

Wcs <- scale(W2, center = TRUE, scale = TRUE)
G <- tcrossprod(Wcs)/ncol(Wcs)
```


## Varinace component estimates 
We will use the [rrBLUP](https://cran.r-project.org/web/packages/rrBLUP/index.html) R package.
```{r, eval=FALSE}
library(rrBLUP)
?mixed.solve
fit <-  mixed.solve(y = y, K = G)

fit$beta # BLUE (intercept)

fit$Vu # additive genetic variance

fit$Ve # residual variance

lambda <- fit$Ve / fit$Vu # lambda
```


## Weighted least squares
We will estimate the marker effect 1. 
```{r, eval=FALSE}
Wj <- W[,1, drop=FALSE] # first marker 
W1j <-  cbind(matrix(1, nrow=nrow(Wj)), Wj) # add the intercept to W
colnames(W1j) <- c("Intercept", "Marker1")

ed <- eigen(G) # eigendecomposition
ed.val <- ed$values
ed.vec <- ed$vectors
lambda <- fit$Ve / fit$Vu

ystar <- t(ed.vec) %*% matrix(y)
Wstar <- t(ed.vec) %*% W1j
Omega <- 1/(ed.val + lambda) # weight

MWMstar <- t(Wstar) %*% diag(Omega) %*% Wstar 
MWystar <- t(Wstar) %*% diag(Omega) %*% ystar
```

# Results
```{r, eval=FALSE}
solve(MWMstar) %*% MWystar # manually computed 

fitW <- lm(ystar~ -1 + Wstar, weights = Omega) # use the lm() function 
summary(fitW)$coef
```


# Two-step approach
Compare the results from weighted least squares GWAS with the two-step approach. 
```{r eval=FALSE, echo=TRUE}
Z <- diag(nrow(G))
I <- diag(nrow(G))
Ginv <- solve(G)
V <- G*fit$Vu + I*fit$Ve
Vinv <- solve(V)

a <- solve(t(W1j) %*% Vinv %*% W1j) %*% t(W1j) %*% Vinv %*% y # # the intercept and the first marker effect
g <- solve(I + Ginv*lambda) %*% (y - X %*% b) # BLUP (genetic values)

a # the intercept and the first marker effect
head(g)
tail(g)
```


# Mixed model equations (MME)
Compare the results from weighted least squares GWAS with the MME. 
```{r eval=FALSE, echo=TRUE}
WtW <- t(W1j) %*% W1j # X'X
WtZ <- t(W1j) %*% Z # X'Z
ZtW <- t(Z) %*% W1j # Z'X
ZtZ <- t(Z) %*% Z # Z'Z
Wty <- t(W1j) %*% y # X'y
Zty <- t(Z) %*% y # Z'y

LHS1 <- cbind(WtW, WtZ)
LHS2 <- cbind(ZtW, ZtZ + Ginv * lambda)
LHS <- rbind(LHS1, LHS2) # left hand side of MME
RHS <- rbind(Wty, Zty) # right hand side of MME

sol <- solve(LHS, RHS) # solve for BLUE and BLUP

# intercept
sol[1, ]

# effect of marker 1
sol[2, ]

# BLUP (genetic values)
sol[-c(1,2)]
```
