---
title: "Estimating genomic heritability from Bayesian marker-based regression models"
author: "Gota Morota"
date: "April 27, 2020"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
This excersie shows how to estimate genomic heritability from Bayesian merker-based regression. Recall that genomic heritability is given by $h^2_g = \frac{\sigma^2_g}{\sigma^2_g + \sigma^2_e}$, where $\sigma^2_g$ is the estimate of genomic variance and $\sigma^2_e$ is the estimate of residual variance. 

# Data 
We will use the `mice` data available in the [BGLR](https://cran.r-project.org/web/packages/BGLR/index.html) R package. 
```{r, eval=FALSE, echo=TRUE}
library(BGLR)
data(mice)
y <- mice.pheno$Obesity.BMI
W <- mice.X
dim(W)
```

## Quality control
We will perform quality control by removing markers with MAF < 0.1. 
```{r, eval=FALSE, echo=TRUE}
freq <- colSums(W) / (2*nrow(W))
maf <- ifelse(freq > 0.5, 1-freq, freq)
maf.index <- which(maf < 0.1)
length(maf.index)
W <- W[, -maf.index]
freq <- freq[-maf.index]
dim(W) # 1814 9340
``` 

## Standarization of marker matrix
We will create type marker matrices. `Wc`: a centered marker matrix. `Wcs`: a centered and scaled marker matrix.
```{r, eval=FALSE, echo=TRUE}
Wc <- scale(W, center = TRUE, scale = FALSE)

Wcs <- scale(W, center = TRUE, scale = TRUE)
```


## Genomic relationship matrix
We will create two types of VanRaden's [genomic relationship matrices](https://doi.org/10.3168/jds.2007-0980): `G1`: VanRaden's first genomic relationship matrix. `G2`: VanRaden's second genomic relationship matrix. 
```{r, eval=FALSE, echo=TRUE}
G1 <- tcrossprod(Wc) / sum(2*freq*(1-freq))

G2 <- tcrossprod(Wcs) / ncol(Wcs)
```

# GBLUP 
Genomic best linear unbiased prediction (GBLUP) is given by 
\begin{align*}
    \mathbf{y} &= \mathbf{Xb} + \mathbf{Zg} + \mathbf{e} \\
    \mathbf{g} &\sim N(0, \mathbf{G}\sigma^2_g) \\
    \mathbf{e} &\sim N(0, \mathbf{I}\sigma^2_e) 
\end{align*}
where $\mathbf{X}$ and $\mathbf{Z}$ are the incidence matrices for fixed and random effects, respectively, and  $\mathbf{b}$ and $\mathbf{g}$ are BLUE of fixed effect and BLUP of genetic values, respectively. Also, $\sigma^2_g$ is the additive genetic variance and $\sigma^2_e$ is the residual variance. By using the restricted maximum likelihood ([REML](https://en.wikipedia.org/wiki/Restricted_maximum_likelihood)), we can obtain the estimate of genomic heritability as

\begin{align*}
h^2_g = \frac{\sigma^2_g}{\sigma^2_g + \sigma^2_e}
\end{align*}

## GBLUP with G1
Use `mixed.solve` function in the [rrBLUP](https://cran.r-project.org/web/packages/rrBLUP/index.html) R package. We will first estimate genomic heritability using GBLUP coupled with the `G1` matrix. 
```{r eval=FALSE, echo=TRUE}
library(rrBLUP)
fitG1 <- mixed.solve(y = y, K=G1)
# additive genetic variance
fitG1$Vu
# residual variance
fitG1$Ve
# genomic h2
fitG1$Vu / (fitG1$Vu + fitG1$Ve) # 0.195
```
The estimate of heritability from the G1-based GBLUP is 0.195. 


## GBLUP with G2
We will estimate genomic heritability using GBLUP again, but this time with the `G2` matrix. 
```{r eval=FALSE, echo=TRUE}
fitG2 <- mixed.solve(y = y, K=G2)
# additive genetic variance
fitG2$Vu
# residual variance
fitG2$Ve
# genomic h2
fitG2$Vu / (fitG2$Vu + fitG2$Ve) # 0.203
```
The estimate of heritability from the G2-based GBLUP is 0.203, which is slightly different from the one from the G1. 


# RR-BLUP
In ridge regression BLUP (RR-BLUP), we treat the markers as random and fit the whole markers simultaneously. Note that in RR-BLUP, we regress phenotype on the marker matrix instead of regressing on the $G$ matrix as in GBLUP. 

\begin{align*}
\mathbf{y} &= \mathbf{Xb} + \mathbf{Wa} + \mathbf{e} \\
\mathbf{a} & \sim N(0, \mathbf{I}\sigma^2_a) \\
\mathbf{e} & \sim N(0, \mathbf{I}\sigma^2_e)
\end{align*}

- $\mathbf{X}$: incidence matrix of fixed effect
- $\mathbf{b}$: vector of fixed effect
- $\mathbf{W}$: marker matrix
- $\mathbf{a}$: vector of marker effect 
- $\sigma^2_a$: marker genetic variance
- $\mathbf{e}$: vector of residual
- $\sigma^2_a$: residual variance

The marker genetic variance $\sigma^2_a$ is not genetic variance $\sigma^2_g$, which is required to calculate genomic heritability. Therefore, we have to indirectly estimate $\sigma^2_g$ to obtain the estimate of genomic heritability. 

## RR-BLUP with Wc
In RR-BLUP, we estimate genomic heritability by regressing phenotype on markre covariates. We first estimate marker genetic variance $\sigma^2_a$ in RR-BLUP. Then we estimate genetic variance $\sigma^2_g$ from  $\sigma^2_a$  by using the formula $\sigma^2_g = \sum^m_{j=1}(2p_j(1-p_j)) \sigma^2_a$, where $p_j$ is the $j$th allele frequency of reference allele. This formula holds when the marker matrix is centered (`Wc`). 
```{r eval=FALSE, echo=TRUE}
library(rrBLUP)
fitWc <- mixed.solve(y = y, Z=Wc)
# marker additive genetic variance
fitWc$Vu
# residual variance
fitWc$Ve
# marker additive genetic effects
head(fitWc$u)
# additive genetic variance
sigma2gWc <- sum(2*freq*(1-freq)) * fitWc$Vu 
# genomic h2
sigma2gWc / (sigma2gWc + fitWc$Ve) # 0.195
```
We can confirm that the estimate of genomic heritability is equivalent to the one from G1-based GBLUP.



## RR-BLUP with Wcs
Similar to the above, we first estimate marker genetic variance $\sigma^2_a$ in RR-BLUP. Then we estimate genetic variance $\sigma^2_g$ from  $\sigma^2_a$  by using the formula $\sigma^2_g = m \sigma^2_a$, where $m$ is the number of markers. This formula holds when the marker matrix centered and scaled (`Wcs`). 
```{r eval=FALSE, echo=TRUE}
library(rrBLUP)
fitWcs <- mixed.solve(y = y, Z=Wcs)
# marker additive genetic variance
fitWcs$Vu
# residual variance
fitWcs$Ve
# marker additive genetic effects
head(fitWcs$u)
# additive genetic variance
sigma2gWcs <- ncol(Wcs) * fitWcs$Vu 
# genomic h2
sigma2gWcs / (sigma2gWcs + fitWcs$Ve) # 0.203
```
We can confirm that the estimate of genomic heritability is equivalent to the one from G2-based GBLUP.


# Bayesian ridge regression
Bayesian ridge regression is equivalent to assigning a Gaussian prior to the marker effect. 
\begin{align*}
    \mathbf{y} = \sum^m_{j=1} \mathbf{W}_{j} a_j + \epsilon
    \rightarrow \mathbf{y}| a_j, \sigma^2_e \sim N(\mathbf{W}_{j} a_j, \sigma^2_e)
\end{align*}
  
$p(\mathbf{a}| \mathbf{y}, \sigma^2_e, \sigma^2_{a}) \propto p(\mathbf{y}|\mathbf{a}, \sigma^2_e) \cdot p(\mathbf{a}| \sigma^2_{a})$

\begin{align*}
        \text{Prior distributions}
    \begin{cases}
          a_j|\sigma^2_{a} \sim N(0,  \sigma_{a}^{2} ) \\
          \sigma^2_{a} \sim \chi^{-2}(\nu, S)  \\
          \sigma^2_{e} \sim \chi^{-2}(\nu, S)
    \end{cases}
\end{align*}

- $\chi^{-2}(\nu, S) \rightarrow$ scaled inverted chi-square distribution with $\nu$ degrees of freedom  and scale parameter $S$ 

## Bayesian ridge regression with Wc
We can estimate genomic heritability from Bayesian ridge regression with Wc as the same way as in RR-BLUP with Wc using the following relationship between $\sigma^2_g$ and $\sigma^2_a$. 

\[
\sigma^2_g = \sum^m_{j=1}(2p_j(1-p_j)) \sigma^2_a
\]

```{r eval=FALSE, echo=TRUE}
ETA <- list(MRK=list(X = Wc, model = "BRR"))
fitBRRWc <- BGLR(y = y, ETA = ETA, nIter = 3000, burnIn = 500, verbose = TRUE)

# marker genetic variance
fitBRRWc$ETA[[1]]$varB
# residual variance
fitBRRWc$varE
# marker additive genetic effects
fitBRRWc$ETA[[1]]$b
# additive genetic variance
sigma2gBRRWc <- sum(2*freq*(1-freq)) * fitBRRWc$ETA[[1]]$varB
# genomic h2
sigma2gWc / (sigma2gWc + fitBRRWc$varE) # 0.198
```
We can see that the estimate of genomic heritability from Bayesian ridge regression with Wc is very simlar to the one from RR-BLUP with Wc. 


## Bayesian ridge regression with Wcs
We can estimate genomic heritability from Bayesian ridge regression with Wcs as the same way as in RR-BLUP with Wcs by using the following relationship between $\sigma^2_g$ and $\sigma^2_a$. 

\[
\sigma^2_g = m \sigma^2_a
\]

```{r eval=FALSE, echo=TRUE}
set.seed(100)
ETA <- list(MRK=list(X = Wcs, model = "BRR"))
fitBRRWcs <- BGLR(y = y, ETA = ETA, nIter = 3000, burnIn = 500, verbose = TRUE)

# marker genetic variance
fitBRRWcs$ETA[[1]]$varB
# residual variance
fitBRRWcs$varE
# marker additive genetic effects
head(fitBRRWcs$ETA[[1]]$b)
# additive genetic variance
sigma2gBRRWcs <- ncol(Wcs) * fitBRRWcs$ETA[[1]]$varB
# genomic h2
sigma2gWcs / (sigma2gWcs + fitBRRWcs$varE) # 0.205
```
We can see that the estimate of genomic heritability from Bayesian ridge regression with Wcs is very simlar to the one from RR-BLUP with Wcs. 


# BayesC
BayesC is equivalent to assuming two component mixture prior with a point of mass at zero and a Gaussian slab. It does both shrinkage and variable selection.
  \begin{align*}
    \mathbf{y} = \sum^m_{j=1} \mathbf{W}_{j} a_j + \epsilon
    \rightarrow \mathbf{y}| a_j, \sigma^2_e \sim N(\mathbf{W}_{j} a_j, \sigma^2_e)
\end{align*}

$p(\mathbf{a}| \mathbf{y}, \sigma^2_e, \sigma^2_{a}) \propto p(\mathbf{y}|\mathbf{a}, \sigma^2_e) \cdot p(\mathbf{a}| \sigma^2_{a})$

\begin{align*}
        \text{Prior distributions}
        \begin{cases}
           \begin{cases}
                       a_j = 0  \quad \text{with probability} \quad  \pi  \\
                       a_j|\sigma^2_{a} \sim N(0,  \sigma_{a}^{2} )  \quad \text{with probability} \quad (1 - \pi)
           \end{cases} \\
           \sigma^2_{a} \sim  \chi^{-2}(\nu, S)  \\
          \sigma^2_{e} \sim \chi^{-2}(\nu, S)
    \end{cases}
\end{align*}

- $\pi$: probability that markers have no effect

- $\chi^{-2}(\nu, S) \rightarrow$ scaled inverted chi-square distribution with $\nu$ degrees of freedom  and scale parameter $S$ 

Unlike Bayesian ridge regression, the relationship between $\sigma^2_g$ and $\sigma^2_a$ is not clear in BayesC. Theoretically, it is possible to derive a sample of the genetic variance from the sample variance of the genetic values at each iteration of the Gibbs sampler as shown in this [paper](https://doi.org/10.1534/genetics.117.300271). For $s$th Gibbs sample, we calculate 

\begin{align*}
\sigma^{2(s)}_g = \frac{\sum_{i=1}^{n} (g^{(s)}_i - \bar{g}^{(s)})^2}{n}
\end{align*}
where $\bar{g}^{(s)} = \frac{\sum^n_{i=1} g_i^{(s)}}{n}$ is the average genetic value and $n$ is the number of individuals. Then, compute genomic heritability at $s$th iteration as 
\begin{align*}
h^{2(s)}_g = \frac{\sigma^{2(s)}_g}{\sigma^{2(s)}_g+ \sigma^{2(s)}_e}
\end{align*}

Once Gibbs sampling is over, compute the posterior mean of the genomic heritability.

## BayesC with Wcs
The following code estimates the genomic heritability from BayesC with Wcs.
```{r eval=FALSE, echo=TRUE}
set.seed(100)
ETA <- list(MRK=list(X = Wcs, model = "BayesC", saveEffects=TRUE))
fitBayesCWcs <- BGLR(y = y, ETA = ETA, nIter = 6000, burnIn = 1000, verbose = TRUE)

# marker genetic variance
fitBayesCWcs$ETA[[1]]$varB
# residual variance
fitBayesCWcs$varE
# marker additive genetic effects
head(fitBayesCWcs$ETA[[1]]$b)

# read output files
B=readBinMat('ETA_MRK_b.bin')
dim(B)
E <- scan("varE.dat")
E <- E[-c(1:fitBayesCWcs$burnIn / fitBayesCWcs$thin)]
length(E)

# compute genomic heritability at each iteration of the Gibbs sampler 
h2vec <- array()
for (s in 1:nrow(B)){
  
  ghat <- Wcs %*% B[s,]
  gbar <- mean(Wcs %*% B[s,])
  sigma2g <- sum( (ghat - gbar)^2 ) / (nrow(Wcs) )
  h2vec[s] <- sigma2g / E[s]

  }

# posterior mean of h2
mean(h2vec) # 0.272
```
The estimate of genomic heritability from BayesC is larger than those of GBLUP, RR-BLUP, and Bayesian ridge regression. 

# BayesB
BayesB is equivalent to assuming two component mixture prior with a point of mass at zero and a sclaed-t distribution slab. It does both shrinkage and variable selection.
 \begin{align*}
    \mathbf{y} = \sum^m_{j=1} \mathbf{W}_{j} a_j + \epsilon
    \rightarrow \mathbf{y}| a_j, \sigma^2_e \sim N(\mathbf{W}_{j} a_j, \sigma^2_e)
\end{align*}

$p(\mathbf{a}| \mathbf{y}, \sigma^2_e, \sigma^2_{a}) \propto p(\mathbf{y}|\mathbf{a}, \sigma^2_e) \cdot p(\mathbf{a}| \sigma^2_{a})$

\begin{align*}
        \text{Prior distributions}
        \begin{cases}
           \begin{cases}
                       a_j =  0  \quad \text{with probability} \quad  \pi  \\
                       a_j|\sigma^2_{a_j} \sim N(0,  \sigma_{a_j}^{2} )  \quad \text{with probability} \quad (1 - \pi)
           \end{cases} \\
           \sigma^2_{a_j} \sim  \chi^{-2}(\nu, S)  \\
          \sigma^2_{e} \sim \chi^{-2}(\nu, S)
    \end{cases}
\end{align*}

- $\pi$: probability that markers have no effect

- $\chi^{-2}(\nu, S) \rightarrow$ scaled inverted chi-square distribution with $\nu$ degrees of freedom  and scale parameter $S$ 

Similar to BayesC, we calculate the genetic variance from the sample variance of the genetic values at each iteration of the Gibbs sampler.

## BayesB with Wcs
The following code estimates the genomic heritability from BayesB with Wcs.
```{r eval=FALSE, echo=TRUE}
set.seed(100)
ETA <- list(MRK=list(X = Wcs, model = "BayesB", saveEffects=TRUE))
fitBayesBWcs <- BGLR(y = y, ETA = ETA, nIter = 6000, burnIn = 1000, verbose = TRUE)

# marker genetic variance
fitBayesBWcs$ETA[[1]]$varB
# residual variance
fitBayesBWcs$varE
# marker additive genetic effects
head(fitBayesBWcs$ETA[[1]]$b)

# read output files
B=readBinMat('ETA_MRK_b.bin')
dim(B)
E <- scan("varE.dat")
E <- E[-c(1:fitBayesBWcs$burnIn / fitBayesBWcs$thin)]
length(E)

# compute genomic heritability at each iteration of the Gibbs sampler 
h2vec <- array()
for (s in 1:nrow(B)){
  
  ghat <- Wcs %*% B[s,]
  gbar <- mean(Wcs %*% B[s,])
  sigma2g <- sum( (ghat - gbar)^2 ) / (nrow(Wcs) )
  h2vec[s] <- sigma2g / E[s]

}

# posterior mean of h2
mean(h2vec) # 0.303
```
The estimate of genomic heritability from BayesB is overestimated compared to those of GBLUP, RR-BLUP, and Bayesian ridge regression. 



