---
title: "Cross-validation for genomic prediction"
author: "Gota Morota"
date: "April 3, 2020"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Data 
We will use the `mice` data available in the [BGLR](https://cran.r-project.org/web/packages/BGLR/index.html) R package. 
```{r, eval=FALSE, echo=TRUE}
library(BGLR)
data(mice)
y <- mice.pheno$Obesity.BMI
W <- mice.X
dim(W)
```

## Quality control
We will perform quality control by removing markers with MAF < 0.1. 
```{r, eval=FALSE, echo=TRUE}
freq <- colSums(W) / (2*nrow(W))
maf <- ifelse(freq > 0.5, 1-freq, freq)
maf.index <- which(maf < 0.1)
length(maf.index)
W <- W[, -maf.index] 
dim(W)
```

## Genomic relationship matrix
We will create a genomic relationship matrix.  
```{r, eval=FALSE, echo=TRUE}
Wcs <- scale(W, center = TRUE, scale = TRUE)
G <- tcrossprod(Wcs) / ncol(Wcs)
```

## RR-BLUP
We will treat the first 1000 individuals as a training set and predict the additive genetic values of the remaining individuals in the testing set. 
```{r eval=FALSE, echo=TRUE}
library(rrBLUP)
n.trn <- 1000
n.tst <- length(y) - 1000
n.tst
y.trn <- y[1:n.trn]
length(y.trn)
y.tst <- y[n.trn+1:n.tst]
length(y.tst)
W.trn <- Wcs[1:n.trn, ]
dim(W.trn)
W.tst <- Wcs[n.trn+1:n.tst, ]
dim(W.tst)

fit.trn <- mixed.solve(y = y.trn, Z = W.trn) # estimate marker effects from the training set
a.trn <- fit.trn$u

y.tst.hat1 <- W.tst %*% a.trn # prediction of the testing set individuals
```

What is the predictive correlation in the testing set?
```{r, eval=FALSE, echo=TRUE}
cor(y.tst, y.tst.hat1)
```




## GBLUP
We will treat the first 1000 individuals as a training set and predict the additive genetic values of the remaining individuals in the testing set. 
```{r eval=FALSE, echo=TRUE}
Gtrn <- G[1:n.trn, 1:n.trn]
dim(Gtrn)
Gtst.trn <- G[n.trn+1:n.tst, 1:n.trn]
dim(Gtst.trn)
Ginv.trn <- solve(Gtrn)

fit.trn2 <- mixed.solve(y = y.trn, K = Gtrn)
g.trn.hat <- fit.trn2$u  # estimate the genetic values of the training set individuals

y.tst.hat2 <- Gtst.trn %*% Ginv.trn %*% g.trn.hat # prediction of the testing set individuals
```


What is the predictive correlation in the testing set?
```{r, eval=FALSE, echo=TRUE}
# prediction
cor(y.tst, y.tst.hat2) 
```


